Forcing for Isabelle

### 2018-12-12 LICS/isabelle.tex
edits variados
ediciones varias

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Isabelle}
Isabelle provides a meta-language called \emph{Pure} that consists of
a fragment of higher order logic. The meta-Boolean type is called
\isatt{prop}. Meta connectives
\isatt{\isasymLongrightarrow} and \isatt{\&\&\&} fulfill the role of
implication and conjunction, and the meta-binder \isatt{\isasymAnd}
corresponds to universal quantification.

On top of \emph{Pure}, theories/object logics can be defined, with
their own types, connectives and rules. Rules can be written  using
meta-implication: ``$P$, $Q$, and $R$ yield $S$'' can be written
\[
P \ \isatt{\isasymLongrightarrow}\ Q\ \isatt{\isasymLongrightarrow}\ R\ \isatt{\isasymLongrightarrow}\ S
\]
(as usual,  \isatt{\isasymLongrightarrow} asociates to the right), and
syntactic sugar is provided to curry the previous as follows:
\[
\isasymlbrakk P; Q; R \isasymrbrakk \ \isatt{\isasymLongrightarrow}\ S.
\]

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Isabelle}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
\subsection{Logics}

Isabelle provides a meta-language called \emph{Pure} that consists of
a fragment of higher order logic. The meta-Boolean type is called
\isatt{prop}. Meta-connectives
\isatt{\isasymLongrightarrow} and \isatt{\&\&\&} fulfill the role of
implication and conjunction, and the meta-binder \isatt{\isasymAnd}
corresponds to universal quantification.

On top of \emph{Pure}, theories/object logics can be defined, with
their own types, connectives and rules. Rules can be written  using
meta-implication: ``$P$, $Q$, and $R$ yield $S$'' can be written
\[
P \ \isatt{\isasymLongrightarrow}\ Q\ \isatt{\isasymLongrightarrow}\ R\ \isatt{\isasymLongrightarrow}\ S
\]
(as usual,  \isatt{\isasymLongrightarrow} asociates to the right), and
syntactic sugar is provided to curry the previous as follows:
\[
\isasymlbrakk P; Q; R \isasymrbrakk \ \isatt{\isasymLongrightarrow}\ S.
\]
A further example is given by induction on the natural numbers
\isatt{nat},
\[
\isasymlbrakk P(0);\ (\textstyle\isasymAnd
x.\ P(x)\ \isasymLongrightarrow\ P(\isatt{succ}(x))) \isasymrbrakk
\ \isasymLongrightarrow\ P(n), 
\]
where we are omitting the ``typing'' assumtions on $n$ and $x$.

@@

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
\subsection{Locales}

Locales \cite{ballarin2010tutorial} provide a neat facility to
encapsulate a context (fixed objects and assumptions on them) that is
to be used in proving several theorems, as in usual mathematical
practice.

In this paper, locales have a further use. The \emph{Fundamental
  Theorems of Forcing} we use talk about a specific map $\forceisa$
from formulas to formulas. The definition of $\forceisa$ is involved
and we will not dwell on this now; but applications of those theorems
do not require to know how it is defined. Therefore, we black-box it
and package everything in a locale that assumes that there is such a
map that satisfies the Fundamental Theorems.

### 2018-12-12 LICS/zfc-axioms.tex
ediciones varias

@@

The existence of a ctm of $\ZFC$ can be proved from the existence of a
model  $\lb N,E\rb$ such that the relation $E$ is well founded. The
L\"owenheim-Skolem 
Theorem ensures that there is an countable elementary submodel 
$\lb N',E\restr N'\rb\preccurlyeq  \lb N,E\rb$ which must also be
well founded; then the 
Mostowski collapsing function \cite[Def.~I.9.31]{kunen2011set} sends $\lb
N',E\restr N'\rb$ 
isomorphically to an $\lb M,\in\rb$ where $M$ is transitive.

>>>>

The axioms of Zermelo and Fraenkel with Choice ($\ZFC$) are a
countably infinite list of first-order sentences.
\medskip
\fbox{\em \dots\ discussion of axioms \dots }
\medskip

We chose a presentation of the $\ZFC$ axioms that would be most
compatible with the development of relativization by Paulson:
\begin{isabelle}
\isacommand{locale}\isamarkupfalse%
\ M{\isacharunderscore}ZF\ {\isacharequal}\ \isanewline
\ \isakeyword{fixes}\ M\ \isanewline
\ \isakeyword{assumes}\ \isanewline
\ \ \ \ \ \ upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ Union{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ power{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}power{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ extensionality{\isacharcolon}\ \ \ {\isachardoublequoteopen}extensionality{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ foundation{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ infinity{\isacharunderscore}ax{\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ separation{\isacharunderscore}ax{\isacharcolon}\isanewline
\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymrbrakk}\isanewline\ \ \ \ {\isasymLongrightarrow}\isanewline
\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{and}\ replacement{\isacharunderscore}ax{\isacharcolon}\isanewline 
\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{3}}\ {\isasymrbrakk}\isanewline
\ \ \ \ {\isasymLongrightarrow}\isanewline
\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}\isanewline
\ \ \ \ {\isasymlambda}x\ y{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}y{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

@@

The existence of a ctm of $\ZFC$ can be proved from the existence of a
model  $\lb N,E\rb$ such that the relation $E$ is well founded. The
L\"owenheim-Skolem 
Theorem ensures that there is an countable elementary submodel 
$\lb N',E\restr N'\rb\preccurlyeq  \lb N,E\rb$ which must also be
well founded; then the 
Mostowski collapsing function \cite[Def.~I.9.31]{kunen2011set} sends $\lb
N',E\restr N'\rb$ 
isomorphically to some $\lb M,\in\rb$ with  $M$ transitive.
